---
title: "SP500 forecast"
author: "Moises Evangelista"
date: "Prepared `r format(Sys.time(), '%B %d, %Y')`"
output: 
  beamer_presentation:
    theme: "AnnArbor"
    fonttheme: "serif"
    fig_width: 7
    fig_height: 6
    fig_caption: true
    fig_crop: false
---

## Introduction

The purpose of this analysis is to determine if the SP500 Index (Index) is overpriced, as of August 2015, and complete some forecasting analysis.

A forecast of the Index. The Index data was downloaded from [Yahoo Finance](http://finance.yahoo.com/q?s=%5EGSPC) in August after if had significant drops.  Consumer Price Index data was used to deflate the Index's historical price.  This data can be downloaded from [here](http://www.bls.gov/data/).

In conclusion, it seems that the Index is currently overpriced.


## Analysis steps

- The data was downloaded
- Historical data was visualized
- Historical data adjusted by the CPI and then was visualized
- Forecasts were completed


```{r echo=FALSE, message = FALSE, include=FALSE}

rm(list = ls(all = TRUE)) #start with empty workspace
startTime <- Sys.time()


library(knitr)
library(RCurl)
library(dplyr)
library(data.table)
library(ggplot2)
library(scales)
library(grid)
library(tidyr)
library(forecast)
library(reshape2)
library(cv.ts)
library(cowplot)
# library(devtools)
# install_github('sinhrks/ggfortify')
library(ggfortify)

setwd("~/GitHub/SP500-Forecast/") # set the working directory
list.files() # see whats there

```

```{r import data, echo = FALSE, message = FALSE, cache=TRUE, warning=FALSE, include=FALSE}

linkSp500 <- "http://real-chart.finance.yahoo.com/table.csv?s=%5EGSPC&a=00&b=3&c=1950&d=07&e=26&f=2016&g=m&ignore=.csv"

SP500MonthlyData <- fread(linkSp500,
                          data.table = FALSE) %>%
  mutate(Date = as.Date(Date, format = "%Y-%m-%d"),
         Year = as.integer(format(Date, "%Y")),
         Period = as.integer(format(Date, "%m")))

# import csv data. 
# data is from here, but saved locally
# http://data.bls.gov/pdq/SurveyOutputServlet or http://data.bls.gov/cgi-bin/surveymost
# data info is here http://www.bls.gov/cpi/#tables

CpiLink <- "http://download.bls.gov/pub/time.series/cu/cu.data.1.AllItems"

ConsumerPriceIndexAllUrbanConsumersMonths <- fread(CpiLink,
                                                   data.table = FALSE) %>% 
  filter(  grepl('M', period)) %>%
  filter(year >= 1950,
         series_id == "CUUR0000SA0") %>%
  mutate(period = as.integer(gsub("M","", period)))

comboData <- left_join(SP500MonthlyData, ConsumerPriceIndexAllUrbanConsumersMonths,
                       by = c("Year" = "year",
                              "Period" = "period")) %>%
  mutate(OpenHigh = (Open/value)*100,
         OpenHighLog  = log10(OpenHigh)) %>%
  filter(!is.na(OpenHighLog))

summary(comboData)
str(comboData)


```

## SP500 Open Price by Month 

```{r, echo =FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=6, fig.cap="Open market price trend Jan 2014 - Jul 2015"}

options(digits = 2)

# get the end and start dates of the data
end <- max(comboData$Date)
start <- min(comboData$Date)

endRecent <- max(comboData$Date)
startRecent <- as.Date('2015-01-01')

# code is from here
# https://learnr.wordpress.com/2009/05/08/ggplot2-plot-inside-a-plot/

theme_white <- function() {
  theme_update(panel.background = element_blank(),
               panel.grid.major = element_blank())
}
theme_set(theme_bw())

theme_white()

p <- ggplot(comboData, aes(Date, Open)) + labs(x = NULL,
                                                    y = ("Open Price")) 

mainplot <- p %+%  
  geom_line(data = comboData,
            colour = "grey",
            size = 0.53) + 
  geom_line(stat = "smooth",method = "loess", color = "blue",
            size = .5,
            alpha = 0.5) +
  geom_ribbon(stat = 'smooth', method = "loess", 
              se = TRUE,
              alpha = 0.1, 
              aes(color = NULL)) +
  coord_cartesian(xlim = as.Date(c("2014-01-01", "2016-05-02"))) +
  scale_y_continuous(labels = comma_format(),
                     # limits=c(1500, 2200)
  ) +
  theme(axis.title.x = element_text(size = 8, 
                                    angle = 00),
        axis.text.x = element_text(colour = "black", 
                                   size = 5, 
                                   angle = 00, 
                                   vjust = .5),
        axis.title.y = element_text(size = 8, 
                                    angle = 90),
        axis.text.y = element_text(colour = "black", 
                                   size = 5, 
                                   angle = 00, 
                                   vjust = .5),
        plot.margin = unit(c(0,0,0,0), "cm"))

p1 <- p + geom_rect(aes(xmin = as.Date('2014-01-01'), 
                        xmax = max(comboData$Date),
                        ymin = min(Open, na.rm = TRUE), 
                        ymax = max(Open, na.rm = TRUE)),
                    fill = alpha("lightblue", 0.2)) + 
  theme(axis.title.x = element_blank(),
        # axis.text.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        # panel.border = element_blank(),
        panel.background = element_blank(),
        plot.background = element_blank(),
        panel.grid = element_blank(),
        plot.margin = unit(c(0,0,-1,-1), "cm"))  #T, R, B, L


subplot <- p1 + geom_line( colour = "grey",
                           size = 0.53) + 
  geom_line(stat = "smooth",method = "loess", color = "blue",
            size = .5,
            alpha = 0.5) +
  geom_ribbon(stat = 'smooth', method = "loess", 
              se = TRUE,
              alpha = 0.1, 
              aes(color = NULL)) +
  geom_text(data = data.frame(), aes(as.Date('1980-01-01'),
                                     600,
                                     label = "SP500 Open market price \n1950-2016"),
            size = 2) # +
# ggtitle("SP500 Open market price 1950-2015") +
#theme(plot.title = element_text(lineheight=.8, 
#                                size = 6,
#                               face="bold"))

vp <- viewport(width = 0.5, 
               height = 0.4, 
               x = .9,
               y = unit(3, 
                        "lines"), 
               just = c("right",
                        "bottom"))

full <- function() {
  print(mainplot)
  theme_set(theme_bw(base_size = 8))
  theme_white()
  print(subplot, vp = vp)
  theme_set(theme_bw())
}

full()

# write.table(df, 
#            file = "df.txt",  sep="\t", 
#            row.names = FALSE, na = "")

```

## SP500 Open Price adjusted with CPI by Month 

```{r, echo =FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=6, fig.cap="CPI adjusted open market price trend Jan 2014 - May 2016"}

theme_white()

p <- ggplot(comboData, aes(Date, OpenHigh)) + 
  labs(x = NULL,
       y = ("Open price adjusted with consumer price index (CPI)")) 

mainplot <- p %+%  
  geom_line(data = comboData,
            colour = "grey",
            size = 0.53) + 
  geom_line(stat = "smooth",method = "loess", color = "blue",
            size = .5,
            alpha = 0.5) +
  geom_ribbon(stat = 'smooth', method = "loess", 
              se = TRUE,
              alpha = 0.1, 
              aes(color = NULL)) +
  geom_line(stat = "smooth",method = "lm", color = "red",
            size = .5,
            alpha = 0.5) +
  geom_ribbon(stat = 'smooth', method = "lm", 
              se = TRUE,
              alpha = 0.1, 
              aes(color = NULL)) +
  coord_cartesian(xlim = as.Date(c("2014-01-01", "2016-05-02"))) +
  scale_y_continuous(labels = comma_format(),
                     # limits=c(1500, 2200)
  ) +
  theme(axis.title.x = element_text(size = 8, 
                                    angle = 00),
        axis.text.x = element_text(colour = "black", 
                                   size = 5, 
                                   angle = 00, 
                                   vjust = .5),
        axis.title.y = element_text(size = 8, 
                                    angle = 90),
        axis.text.y = element_text(colour = "black", 
                                   size = 5, 
                                   angle = 00, 
                                   vjust = .5),
        plot.margin = unit(c(0,0,0,0), "cm"))

p1 <- p + geom_rect(aes(xmin = as.Date('2014-01-01'), 
                        xmax = max(comboData$Date),
                        ymin = min(OpenHigh, na.rm = TRUE), 
                        ymax = max(OpenHigh, na.rm = TRUE)),
                    fill = alpha("lightblue", 0.2)) + 
  theme(axis.title.x = element_blank(),
        # axis.text.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        # panel.border = element_blank(),
        panel.background = element_blank(),
        plot.background = element_blank(),
        panel.grid = element_blank(),
        plot.margin = unit(c(0,0,-1,-1), "cm"))  #T, R, B, L

subplot <- p1 + geom_line( colour = "grey",
                           size = 0.53) + 
  geom_line(stat = "smooth",method = "loess", color = "blue",
            size = .5,
            alpha = 0.5) +
  geom_ribbon(stat = 'smooth', method = "loess", 
              se = TRUE,
              alpha = 0.1, 
              aes(color = NULL)) +
  geom_line(stat = "smooth",method = "lm", color = "red",
            size = .5,
            alpha = 0.5) +
  geom_ribbon(stat = 'smooth', method = "lm", 
              se = TRUE,
              alpha = 0.1, 
              aes(color = NULL)) +
  geom_text(data = data.frame(), 
            aes(as.Date('1985-01-01'), 700, 
                label = "SP500 Open market price \n1950-2016"),
            size = 2) + 
  geom_text(data = data.frame(), 
            aes(as.Date('2000-03-01'), 390, 
                label = "Local \nsmoothing"),
            color = "blue",
            size = 2)  + 
  geom_text(data = data.frame(), 
            aes(as.Date('1968-01-01'), 100, 
                label = "Linear \nsmoothing"),
            color = "red",
            size = 2)

vp <- viewport(width = 0.5, 
               height = .45, 
               x = .9,
               y = unit(3, 
                        "lines"), 
               just = c("right",
                        "bottom"))

full()

```

## Forecast graphs

```{r, echo =FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=6, fig.cap="Forecasts"}
# extract open price from the data
# first need to resort the data by date

comboDataOpenPrice <- comboData %>% arrange(Date) %>%
  subset( Date >='1995-01-01',
                               select = c(Open, Date))

dates <- comboData %>% arrange(Date) %>%
  subset( Date >='1995-01-01',
                               select = c( Date)) %>%
      summarise(minYear = min(Date),
              maxYear = max(Date)) %>%
    gather(key, value) %>%
    mutate(YearValue = as.numeric(format(value,'%Y')),
           MonthValue = as.numeric(format(value,'%m')))
  
  MinYear <- dates[1, 3] # row 1 col 3
  MinMonth <- dates[1, 4] # row 1 col 4
  MaxYear <- dates[2, 3]
  MaxMonth <- dates[2, 4]

# make the count a time series object
tsComboDataOpenPrice <- ts(as.vector(comboDataOpenPrice$Open), 
                     start = c(MinYear,
                               MinMonth), 
                     end = c(MaxYear, MaxMonth), frequency = 12)


# now complete a foreacast
# some code is from here
# http://stats.stackexchange.com/questions/140163/timeseries-analysis-procedure-and-methods-using-r


```

```{r, cache=TRUE, echo=FALSE, message=FALSE, warning=FALSE}

x <- tsComboDataOpenPrice

mod_arima = auto.arima(x, ic = 'aicc', stepwise = FALSE)

mod_exponential = ets(x, ic = 'aicc', restrict = FALSE)

mod_neural = nnetar(x, p = 12, size = 25)

mod_tbats = tbats(x, use.parallel = FALSE)

mod_bats = bats(x, use.parallel = FALSE)


mod_stl = stlm(x, s.window=12, ic='aicc', robust=TRUE, method='ets')
 
 
mod_sts = StructTS(x)

mod_holdWinters = HoltWinters(x, beta = FALSE, gamma = FALSE)

mod_linear = tslm(x ~ trend + season)

```

```{r plot some graphs, cache=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=4, eval=FALSE}

par(mar = rep(2, 4),
    cex.axis = .5,
     las=1,
    mfrow = c(4,2))

plot(forecast(mod_arima, 12), include = 120, cex.main = .8)

plot(forecast(mod_exponential, 12), include = 120, cex.main = .8)

plot(forecast(mod_neural, 12), include = 120, cex.main = .8)

# new plots

plot(forecast(mod_tbats, 12), include = 120, cex.main = .8)

plot(forecast(mod_bats, 12), include = 120, cex.main = .8)

# plot(forecast(mod_sts, 12), include=120)

plot(forecast(mod_holdWinters, 12), include = 120, cex.main = .8)

plot(forecast(mod_linear, 12), include = 120, cex.main = .8)

```


```{r newplots, fig.width= 9, fig.height = 4, fig.cap= "newplots.", echo=FALSE, message = FALSE, warning = FALSE, include=TRUE} 


d.arima <- auto.arima(tsComboDataOpenPrice)

d.forecast <- forecast(d.arima, level = c(95), h = 24)

plotForecast <- autoplot(d.forecast) +
  geom_line( aes(y = value),
            colour = "grey",
                           size = 0.5) +
    geom_line( aes(y = Fitted),
            colour = "blue",
                           size = 0.5)

plotMod_arima <- autoplot(forecast(mod_arima, 24)) +
  geom_line( aes(y = value),
            colour = "grey",
                           size = 0.5) +
    geom_line( aes(y = Fitted),
            colour = "blue",
                           size = 0.5)

plotMod_exponential <- autoplot(forecast(mod_exponential, 24)) +
  geom_line( aes(y = value),
            colour = "grey",
                           size = 0.5) +
    geom_line( aes(y = Fitted),
            colour = "blue",
                           size = 0.5)

# this one is different

mod_neural <- forecast(mod_neural, 24)

mod_neuralDF <- data.frame(Value = as.matrix(mod_neural$mean), 
                           date = time(mod_neural$mean))

mod_neuralDFfitted <- data.frame(Value = as.matrix(mod_neural$fitted),
                                 date = time(mod_neural$fitted))

actual <- data.frame(Value = as.matrix(x), 
                     date = time(x))

plotMod_neural <- ggplot() +
    geom_line(data = actual, aes(x = date, y = Value),
             colour = "grey",
                            size = 0.5) +
   geom_line(data = mod_neuralDF, aes(x = date, y = Value),
             colour = "blue",
                            size = 0.5) +
     geom_line(data = mod_neuralDFfitted, aes(x = date, y = Value),
             colour = "blue",
                            size = 0.5)


plotMod_mod_stl <- autoplot(forecast(mod_stl, 24)) +
  geom_line( aes(y = value),
            colour = "grey",
                           size = 0.5) +
    geom_line( aes(y = Fitted),
            colour = "blue",
                           size = 0.5)

plotMod_mod_sts <- autoplot(forecast(mod_sts, 24)) +
  geom_line( aes(y = value),
            colour = "grey",
                           size = 0.5) +
    geom_line( aes(y = Fitted),
            colour = "blue",
                           size = 0.5)

plotMod_tbats <- autoplot(forecast(mod_tbats, 24)) +
  geom_line( aes(y = value),
            colour = "grey",
                           size = 0.5) +
    geom_line( aes(y = Fitted),
            colour = "blue",
                           size = 0.5)

plotMod_bats <- autoplot(forecast(mod_bats, 24)) +
  geom_line( aes(y = value),
            colour = "grey",
                           size = 0.5) +
    geom_line( aes(y = Fitted),
            colour = "blue",
                           size = 0.5)

plotMod_holdWinters <- autoplot(forecast(mod_holdWinters, 24)) +
  geom_line( aes(y = value),
            colour = "grey",
                           size = 0.5) +
    geom_line( aes(y = Fitted),
            colour = "blue",
                           size = 0.5)

plotMod_linear <- autoplot(forecast(mod_linear, 24)) +
  geom_line( aes(y = value),
            colour = "grey",
                           size = 0.5) +
    geom_line( aes(y = Fitted),
            colour = "blue",
                           size = 0.5)


plot_grid(plotForecast,
          plotMod_arima,
          plotMod_exponential,
          plotMod_tbats,
          plotMod_bats,
          plotMod_neural,
          plotMod_mod_stl,
          plotMod_mod_sts,
          plotMod_holdWinters,
          plotMod_linear,
          labels = c("Forecast",
                     "ARIMA",
                     "Exponential",
                     "TBATS",
                     "BATS",
                     "plotMod_neural",
                     "plotMod_mod_stl",
                     "plotMod_mod_sts",
                     "HoldWinters",
                     "Linear"),
          label_size = 5,
          ncol = 5,
          rel_widths = .5,
          rel_heights = 1)

```

```{r, cache= TRUE, echo=FALSE, warning= FALSE}

# code below is from here
# http://stats.stackexchange.com/questions/140163/timeseries-analysis-procedure-and-methods-using-r

ctrl <- tseriesControl(stepSize = 1, maxHorizon = 12, minObs = 36, fixedWindow = TRUE)

models <- list()

models$arima = cv.ts(
  tsComboDataOpenPrice, auto.arimaForecast, tsControl = ctrl,
  ic = 'aicc', stepwise = FALSE)

models$exp = cv.ts(
  tsComboDataOpenPrice, etsForecast, tsControl = ctrl,
  ic = 'aicc', restrict = FALSE)

models$neural = cv.ts(
  tsComboDataOpenPrice, nnetarForecast, tsControl = ctrl,
  nn_p = 6, size = 5)

# the three models below did not work

# models$tbats = cv.ts(
#   tsComboDataOpenPrice, tbatsForecast, tsControl=ctrl,
#   seasonal.periods=12)

# models$bats = cv.ts(
#  tsComboDataOpenPrice, batsForecast, tsControl=ctrl,
#  seasonal.periods=12)

# models$stl = cv.ts(
#  tsComboDataOpenPrice, stl.Forecast, tsControl=ctrl,
#   s.window=12, ic='aicc', robust=TRUE, method='ets')

models$sts = cv.ts(tsComboDataOpenPrice, stsForecast, tsControl = ctrl)

models$naive = cv.ts(tsComboDataOpenPrice, naiveForecast, tsControl = ctrl)

models$theta = cv.ts(tsComboDataOpenPrice, thetaForecast, tsControl = ctrl)

```
```{r}
# Once we've fit the models, we can compare them by MAPE (cv.ts doesn't yet support MASE):

res_overall <- lapply(models, function(x) x$results[13,-1])

res_overall <- Reduce(rbind, res_overall)

row.names(res_overall) <- names(models)

res_overall <- res_overall[order(res_overall[,'MAPE']),]

res_overall <- round(res_overall, 2)

```

## Forecast table

```{r office capacity, results='asis', message=FALSE, warning=FALSE, echo=FALSE, cache=FALSE, include=TRUE, fig.cap="Forecasts"}


options(digits = 1)

Model <- c("naive",
           "theta",
           "exp",
           "arima",
           "sts",
           "neural")

modelTable <- cbind(Model, res_overall)

library(xtable)

print(xtable(modelTable,
             digits = 1, 
             caption = "Measure each model's accuracy",  
             label ="tblCity", 
             # align = "rl{5cm}r{3cm}r{3cm}r{3cm}r{3cm}r{3cm}"
             ), 
      caption.placement = "bottom", 
      include.rownames = FALSE, 
      # table.placement="t",
       format.args = list(big.mark = ",", decimal.mark = "."),
      comment = FALSE)

```

## Forecast graphs 2

```{r plot some graphs3, cache=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=6}

res <- lapply(models, function(x) x$results$MAPE[1:12])

res <- data.frame(do.call(cbind, res))

res$horizon <- 1:nrow(res)

res <- melt(res, id.var = 'horizon', variable.name = 'model', value.name = 'MAPE')

res$model <- factor(res$model, levels = row.names(res_overall))

ggplot(res, aes(x = horizon, y = MAPE, col = model)) +
  geom_line(size = 1, alpha = 1/2) + theme_bw() +
  theme(legend.position = c(.2, .55),
        legend.background = element_rect(fill="transparent")) +
  scale_color_manual(values = c(
    "#1f78b4",
    "#ff7f00",
    "#33a02c",
    "#6a3d9a",
    "#e31a1c",
    "#b15928" #, "#a6cee3", "#fdbf6f", "#b2df8a")
  ))

```


```{r echo =FALSE}

#get the time
endTime <- Sys.time()
```

## Last Page

\footnotesize
Disclaimer:

The data available in this presentation, including data, tables, numbers, graphics, and text (hereinafter collectively referred to as the "Information"), is provided AS IS and for you to view, access, copy, distribute and otherwise use the Information at your own risk. Do your own due diligence.

The analysis was completed on `r format(Sys.time(), "%a %b %d %X %Y")
` in `r round(difftime(endTime, startTime , units = c( "secs")),0)` seconds. 

