---
title: "SP500 forecast"
author: "Moises Evangelista"
date: "August 26, 2015"
output: 
  beamer_presentation:
    theme: "AnnArbor"
    fonttheme: "serif"
    fig_width: 7
    fig_height: 6
    fig_caption: true
    fig_crop: false
---

## Introduction

The purpose of this analysis is to determine if the SP500 Index (Index) is overpriced, as of August 2015, and complete some forecasting analysis.

A forecast of the Index. The Index data was downloaded from [Yahoo Finance](http://finance.yahoo.com/q?s=%5EGSPC) in August after if had significant drops.  Consumer Price Index data was used to deflate the Index's historical price.  This data can be downloaded from [here](http://www.bls.gov/data/).

In conclusion, it seems that the Index is currently overpriced.


## Analysis steps

- The data was downloaded
- Historical data was visualized
- Historical data adjusted by the CPI and then was visualized
- Forecasts were completed


```{r echo=FALSE, message = FALSE, include=FALSE}

rm(list = ls(all = TRUE)) #start with empty workspace
startTime <- Sys.time()


setwd("~/GitHub/SP500-Forecast/") # set the working directory
list.files() # see whats there
library(knitr)
library(RCurl)

```

```{r import data, echo = FALSE, message = FALSE, cache=TRUE}

x <- getURL("http://real-chart.finance.yahoo.com/table.csv?s=%5EGSPC&a=00&b=3&c=1950&d=07&e=26&f=2015&g=m&ignore=.csv")

SP500MonthlyData <- read.csv(text = x)

# the date is a factor so conver it to date

SP500MonthlyData$DateValid <- as.Date(SP500MonthlyData$Date, format = "%Y-%m-%d")

SP500MonthlyData$Year <- as.integer(format(SP500MonthlyData$DateValid, "%Y"))

SP500MonthlyData$Period <- as.integer(format(SP500MonthlyData$DateValid, "%m"))

# import csv data. 
# data is from here, but saved locally
# http://data.bls.gov/pdq/SurveyOutputServlet or http://data.bls.gov/cgi-bin/surveymost

ConsumerPriceIndexAllUrbanConsumers <- read.table("Consumer Price Index - All Urban Consumers.txt",
                                                  header = TRUE, sep =",")

library(dplyr)

# remove records that are not for month periods or before 1950

ConsumerPriceIndexAllUrbanConsumersMonths <- ConsumerPriceIndexAllUrbanConsumers %>% 
  filter(  grepl('M', Period)) %>%
  filter(Year >= 1950)

# extract the numeric part of the month

ConsumerPriceIndexAllUrbanConsumersMonths$Period <- as.integer(gsub("M","", 
                                                                        ConsumerPriceIndexAllUrbanConsumersMonths$Period))

comboData <- left_join(SP500MonthlyData, ConsumerPriceIndexAllUrbanConsumersMonths)

# create a field with adjusted values

comboData$OpenHigh <- comboData$Open/comboData$Value

# get a log value from the preciding field

comboData$OpenHighLog <- log10(comboData$OpenHigh)

```

## SP500 Open Price by Month 

```{r, echo =FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=6, fig.cap="Open market price trend Jan 2014 - Jul 2015"}

library(ggplot2)
library(scales)
library(grid)

options(digits = 2)

# get the end and start dates of the data
end <- max(comboData$DateValid)
start <- min(comboData$DateValid)

endRecent <- as.Date('2015-07-31')
startRecent <- as.Date('2015-01-01')

# code is from here
# https://learnr.wordpress.com/2009/05/08/ggplot2-plot-inside-a-plot/

theme_white <- function() {
  theme_update(panel.background = element_blank(),
               panel.grid.major = element_blank())
}
theme_set(theme_bw())

theme_white()

p <- ggplot(comboData, aes(DateValid, Open)) + labs(x = NULL,
                                                    y = ("Open Price")) 

mainplot <- p %+%  
  geom_line(data = comboData,
            colour = I("grey"),
            size = 0.8) + 
  geom_smooth(data = comboData, se = FALSE) +
  coord_cartesian(xlim = as.Date(c("2014-01-01", "2015-08-03"))) +
  scale_y_continuous(labels = comma_format(),
                     # limits=c(1500, 2200)
  ) +
  theme(axis.title.x = element_text(size = 8, 
                                    angle = 00),
        axis.text.x = element_text(colour = "black", 
                                   size = 5, 
                                   angle = 00, 
                                   vjust = .5),
        axis.title.y = element_text(size = 8, 
                                    angle = 90),
        axis.text.y = element_text(colour = "black", 
                                   size = 5, 
                                   angle = 00, 
                                   vjust = .5),
        plot.margin = unit(c(0,0,0,0), "cm"))

p1 <- p + geom_rect(aes(xmin = as.Date('2014-01-01'), 
                        xmax = max(comboData$DateValid),
                        ymin = min(Open, na.rm = TRUE), 
                        ymax = max(Open, na.rm = TRUE)),
                    fill = alpha("lightblue", 0.2)) + 
  theme(axis.title.x = element_blank(),
        # axis.text.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        # panel.border = element_blank(),
        panel.background = element_blank(),
        plot.background = element_blank(),
        panel.grid = element_blank(),
        plot.margin = unit(c(0,0,-1,-1), "cm"))  #T, R, B, L

subplot <- p1 + geom_line(colour = I("grey"),
                          size = 0.8) + 
  geom_smooth( ) +
  geom_text(data = data.frame(), aes(as.Date('1980-01-01'), 600, label = "SP500 Open market price \n1950-2015"),
            size = 2) # +
# ggtitle("SP500 Open market price 1950-2015") +
#theme(plot.title = element_text(lineheight=.8, 
#                                size = 6,
#                               face="bold"))

vp <- viewport(width = 0.5, 
               height = 0.4, 
               x = .9,
               y = unit(3, 
                        "lines"), 
               just = c("right",
                        "bottom"))

full <- function() {
  print(mainplot)
  theme_set(theme_bw(base_size = 8))
  theme_white()
  print(subplot, vp = vp)
  theme_set(theme_bw())
}

full()

# write.table(df, 
#            file = "df.txt",  sep="\t", 
#            row.names = FALSE, na = "")

```

## SP500 Open Price adjusted with CPI by Month 

```{r, echo =FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=6, fig.cap="CPI adjusted open market price trend Jan 2014 - Jul 2015"}

theme_white()

p <- ggplot(comboData, aes(DateValid, OpenHigh)) + 
  labs(x = NULL,
       y = ("Open price adjusted with consumer price index (CPI)")) 

mainplot <- p %+%  
  geom_line(data = comboData,
            colour = I("grey"),
            size = 0.8) + 
  geom_smooth(data = comboData, se = FALSE) +
  stat_smooth(method = "lm", color = "red") +
  coord_cartesian(xlim = as.Date(c("2014-01-01", "2015-08-03"))) +
  scale_y_continuous(labels = comma_format(),
                     # limits=c(1500, 2200)
  ) +
  theme(axis.title.x = element_text(size = 8, 
                                    angle = 00),
        axis.text.x = element_text(colour = "black", 
                                   size = 5, 
                                   angle = 00, 
                                   vjust = .5),
        axis.title.y = element_text(size = 8, 
                                    angle = 90),
        axis.text.y = element_text(colour = "black", 
                                   size = 5, 
                                   angle = 00, 
                                   vjust = .5),
        plot.margin = unit(c(0,0,0,0), "cm"))

p1 <- p + geom_rect(aes(xmin = as.Date('2014-01-01'), 
                        xmax = max(comboData$DateValid),
                        ymin = min(OpenHigh, na.rm = TRUE), 
                        ymax = max(OpenHigh, na.rm = TRUE)),
                    fill = alpha("lightblue", 0.2)) + 
  theme(axis.title.x = element_blank(),
        # axis.text.x = element_blank(),
        axis.title.y = element_blank(),
        # axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        # panel.border = element_blank(),
        panel.background = element_blank(),
        plot.background = element_blank(),
        panel.grid = element_blank(),
        plot.margin = unit(c(0,0,-1,-1), "cm"))  #T, R, B, L

subplot <- p1 + geom_line(colour = I("grey"),
                          size = 0.8) + 
  geom_smooth( ) +
  stat_smooth(method = "lm", color = "red") +
  geom_text(data = data.frame(), 
            aes(as.Date('1980-01-01'), 5, 
                label = "SP500 Open market price \n1950-2015"),
            size = 2) + 
  geom_text(data = data.frame(), 
            aes(as.Date('2000-01-01'), 4, 
                label = "Local \nsmoothing"),
            color = "blue",
            size = 2)  + 
  geom_text(data = data.frame(), 
            aes(as.Date('1968-01-01'), 1, 
                label = "Linear smoothing"),
            color = "red",
            size = 2)

vp <- viewport(width = 0.5, 
               height = .5, 
               x = .9,
               y = unit(3, 
                        "lines"), 
               just = c("right",
                        "bottom"))

full()

```

## Forecast graphs

```{r, echo =FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=6, fig.cap="Forecasts"}
# extract open price from the data
# first need to resort the data by date

comboData <- comboData[order(as.Date(comboData$DateValid, format = "%Y-%m-%d")),]

comboDataOpenPrice <- subset(comboData, DateValid >='2007-01-01',
                               select = c(Open))

# make the count a time series object
tsComboDataOpenPrice <- ts(as.vector(comboDataOpenPrice), 
                     start = c(2000,1), end = c(2015, 8), frequency = 12)

# now complete a foreacast
# some code is from here
# http://stats.stackexchange.com/questions/140163/timeseries-analysis-procedure-and-methods-using-r

library(forecast)

```

```{r, cache=TRUE, echo=FALSE, message=FALSE, warning=FALSE}

x <- tsComboDataOpenPrice

mod_arima = auto.arima(x, ic = 'aicc', stepwise = FALSE)

mod_exponential = ets(x, ic = 'aicc', restrict = FALSE)

mod_neural = nnetar(x, p = 12, size = 25)

mod_tbats = tbats(x, use.parallel = FALSE)

mod_bats = bats(x, use.parallel = FALSE)
# mod_stl = stlm(x, s.window=12, ic='aicc', robust=TRUE, method='ets')
# mod_sts = StructTS(x)

mod_holdWinters = HoltWinters(x, beta = FALSE, gamma = FALSE)

mod_linear = tslm(x ~ trend + season)

```

```{r plot some graphs, cache=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=4}

par(mar = rep(2, 4),
    cex.axis = .5,
     las=1,
    mfrow = c(4,2))

plot(forecast(mod_arima, 12), include = 120, cex.main = .8)

plot(forecast(mod_exponential, 12), include = 120, cex.main = .8)

plot(forecast(mod_neural, 12), include = 120, cex.main = .8)

# new plots

plot(forecast(mod_tbats, 12), include = 120, cex.main = .8)

plot(forecast(mod_bats, 12), include = 120, cex.main = .8)

# plot(forecast(mod_sts, 12), include=120)

plot(forecast(mod_holdWinters, 12), include = 120, cex.main = .8)

plot(forecast(mod_linear, 12), include = 120, cex.main = .8)

```


```{r, cache= TRUE}

# code below is from here
# http://stats.stackexchange.com/questions/140163/timeseries-analysis-procedure-and-methods-using-r

library(cv.ts)

ctrl <- tseriesControl(stepSize = 1, maxHorizon = 12, minObs = 36, fixedWindow = TRUE)

models <- list()

models$arima = cv.ts(
  tsComboDataOpenPrice, auto.arimaForecast, tsControl = ctrl,
  ic = 'aicc', stepwise = FALSE)

models$exp = cv.ts(
  tsComboDataOpenPrice, etsForecast, tsControl = ctrl,
  ic = 'aicc', restrict = FALSE)

models$neural = cv.ts(
  tsComboDataOpenPrice, nnetarForecast, tsControl = ctrl,
  nn_p = 6, size = 5)

# the three models below did not work

# models$tbats = cv.ts(
#   tsComboDataOpenPrice, tbatsForecast, tsControl=ctrl,
#   seasonal.periods=12)

# models$bats = cv.ts(
#  tsComboDataOpenPrice, batsForecast, tsControl=ctrl,
#  seasonal.periods=12)

# models$stl = cv.ts(
#  tsComboDataOpenPrice, stl.Forecast, tsControl=ctrl,
#   s.window=12, ic='aicc', robust=TRUE, method='ets')

models$sts = cv.ts(tsComboDataOpenPrice, stsForecast, tsControl = ctrl)

models$naive = cv.ts(tsComboDataOpenPrice, naiveForecast, tsControl = ctrl)

models$theta = cv.ts(tsComboDataOpenPrice, thetaForecast, tsControl = ctrl)

```
```{r}
# Once we've fit the models, we can compare them by MAPE (cv.ts doesn't yet support MASE):

res_overall <- lapply(models, function(x) x$results[13,-1])

res_overall <- Reduce(rbind, res_overall)

row.names(res_overall) <- names(models)

res_overall <- res_overall[order(res_overall[,'MAPE']),]

res_overall <- round(res_overall, 2)

```

## Forecast table

```{r office capacity, results='asis', message=FALSE, warning=FALSE, echo=FALSE, cache=FALSE, include=TRUE, fig.cap="Forecasts"}

options(digits = 1)

Model <- c("naive",
           "theta",
           "exp",
           "arima",
           "sts",
           "neural")

modelTable <- cbind(Model, res_overall)

library(xtable)

print(xtable(modelTable,
             digits = 1, 
             caption = "Measure each model's accuracy",  
             label ="tblCity", 
             # align = "rl{5cm}r{3cm}r{3cm}r{3cm}r{3cm}r{3cm}"
             ), 
      caption.placement = "bottom", 
      include.rownames = FALSE, 
      # table.placement="t",
       format.args = list(big.mark = ",", decimal.mark = "."),
      comment = FALSE)

```

## Forecast graphs 2

```{r plot some graphs3, cache=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=6}

library(reshape2)

library(ggplot2)

res <- lapply(models, function(x) x$results$MAPE[1:12])

res <- data.frame(do.call(cbind, res))

res$horizon <- 1:nrow(res)

res <- melt(res, id.var = 'horizon', variable.name = 'model', value.name = 'MAPE')

res$model <- factor(res$model, levels = row.names(res_overall))

ggplot(res, aes(x = horizon, y = MAPE, col = model)) +
  geom_line(size = 1, alpha = 1/2) + theme_bw() +
  theme(legend.position = c(.2, .55),
        legend.background = element_rect(fill="transparent")) +
  scale_color_manual(values = c(
    "#1f78b4",
    "#ff7f00",
    "#33a02c",
    "#6a3d9a",
    "#e31a1c",
    "#b15928" #, "#a6cee3", "#fdbf6f", "#b2df8a")
  ))

```


```{r echo =FALSE}

#get the time
endTime <- Sys.time()
```

## Last Page

\footnotesize
Disclaimer:

The data available in this presentation, including data, tables, numbers, graphics, and text (hereinafter collectively referred to as the "Information"), is provided AS IS and for you to view, access, copy, distribute and otherwise use the Information at your own risk. Do your own due diligence.

The analysis was completed on `r format(Sys.time(), "%a %b %d %X %Y")
` in `r round(difftime(endTime, startTime , units = c( "secs")),0)` seconds. 

